The \dealii{} Library, Version 9.3

### 2021-05-22 9.3/paper.tex
add my affil
Improve some text. Write about FEPointEvaluation. Add references.

\author[4]{Timo~Heister}

\author[7]{Luca~Heltai}
\affil[7]{SISSA,
   International School for Advanced Studies,
   Via Bonomea 265,
   34136, Trieste, Italy.
   {\texttt{luca.heltai@sissa.it}}}

 \author[8]{Martin~Kronbichler}
 \affil[8]{Institute for Computational Mechanics,
   Technical University of Munich,
   Boltzmannstr.~15, 85748 Garching, Germany.
   {\texttt{kronbichler/munch@lnm.mw.tum.de}}}

\author[9]{Matthias~Maier}
\affil[9]{Department of Mathematics,
  Texas A\&M University,
  3368 TAMU,
  College Station, TX 77845, USA.
  {\texttt{maier@math.tamu.edu}}}

\author[8,10]{Peter Munch}
 \affil[10]{Institute of Materials Research, Materials Mechanics,
 Helmholtz-Zentrum Geesthacht,
 Max-Planck-Str. 1, 21502 Geesthacht, Germany.
   {\texttt{peter.muench@hzg.de}}}

\author[11]{Jean-Paul~Pelteret}
\affil[11]{Independent researcher.
{\texttt{jppelteret@gmail.com}}}

@@

We provide matrix-free support for simplex and mixed meshes as well as for
continuous and discontinuous elements. From the user perspective, the setup has
hardly changed:
%\begin{c++}
%Simplex::QGauss<dim> quad(degree + 1)
%
%MatrixFree matrix_free;
%matrix_free.reinit(mapping, dof_handler, constraints, quad);
%\end{c++}
Instead of providing a 1D quadrature rule, which is extended to higher dimensional
spaces via a tensor product, one provides d-dimensional quadrature rules.
The templated versions of \texttt{FEEvaluation} and \texttt{FEFaceEvaluation} cannot be used: the number of degrees of freedom and the number of quadrature points are determined at runtime. More details can be found in Subsection~\ref{subsec:mf}.

By the time of writing, we do not use internally any advance techniques for evaluating values and
gradients at the quadrature points as we do for tensor-product elements and
instead rely on full interpolation matrices. This is for now acceptable since only low-order elements are
supported.

@@

Until now, \dealii has only supported local smoothing multigrid
algorithms~\citep{ClevengerHeisterKanschatKronbichler2019}; in local
smoothing algorithms, one only applies the smoother to that part of an
adaptively refined mesh that has cells on a given refinement level,
but not to parts of the mesh that are coarser than that level.
The current release now also has support for global coarsening~\citep{becker00} when
using continuous (\texttt{FE\_Q}, \texttt{FE\_SimplexP}) and
discontinuous (\texttt{FE\_DGQ}, \texttt{FE\_SimplexDGP})
elements. Global coarsening builds multigrid levels for the entire
domain, coarsening \textit{all} cells of a triangulation regardless of
how many times they have been refined. In addition, the framework now available in
\dealii{} in not only applicable to geometric coarsening, but can also
perform coarsening by reducing the polynomial degree used, for example
to support $hp$-adaptive meshes where it has been shown that one can
efficiently precondition high-order spaces with lower-order finite
element operators.
Finally, the implementation also supports transfer between continuous
and discontinuous elements by providing \texttt{DoFHandler}s set up
with these kinds of elements; the scheme therefore allows
preconditioning of discontinuous schemes with a continuous
discretization with fewer unknowns, for example.

@@

The transfer operators between two levels has been implemented in the new class \texttt{MGTwoLevelTransfer}, which can be set up via the functions \texttt{MGTwoLevelTransfer::\allowbreak reinit\_\allowbreak geometric\_\allowbreak transfer()} or \texttt{MGTwo\allowbreak LevelTransfer::\allowbreak reinit\_\allowbreak polynomial\_\allowbreak transfer()} for given
\texttt{DoFHandler} and \texttt{AffineConstraint} classes of two levels. The resulting transfer operators
can then be collected in a single
\texttt{MGTransferGlobalCoarsening} object that can be used just as \texttt{MGTransferMatrixFree} and can be passed to the actual \texttt{Multigrid}
algorithm.
Several common operations are encoded in utility
functions in the \texttt{MGTransferGlobalCoarseningTools}
namespace. In practice, matrix-free methods typically require
constructing the diagonal of the system matrix for smoothing
operations, and the full matrix on the coarse level for the coarse-grid solver;
the functions \texttt{create\_diagonal()} and \texttt{create\_matrix()} from
the \texttt{MatrixFreeTools} namespace can be used to this end (see also Subsection~\ref{subsec:mf}).

@@

The classes \texttt{FEEvaluation} and \texttt{FEFaceEvaluation} are highly
templated to reach high performance. In particular, the template parameters include
the polynomial degree of the finite element $k$ and the number of the 1D quadrature points $q$.
For application codes that rely on operators of many different degrees (e.g., because
they use $p$-multigrid or $hp$-algorithms), the many instantiations
necessary and their complexity
imply long compile times.

In the current release, we have improved specializations of these classes that do
not rely on the template parameters $k$ and $q$ (expressed in the code
with ``-1'' and ``0'') -- for example:
\begin{c++}
FEEvaluation<dim, -1, 0, n_components, Number, VectorizedArrayType>
  phi(range, dofhandler_index, quadrature_index, first_selected_component);
\end{c++}
These classes select at runtime -- for common low and medium
polynomial degree/quadrature combinations ($k\le 6$ and $q\in\{ k+1,
k+2, \left\lfloor (3k)/2 \right\rfloor \}$) -- efficient precompiled implementations and default to non-templated
evaluation kernels else.

@@

With release 9.1 \cite{dealII91}, large parts of the $hp$-algorithms in \dealii were parallelized so that
parallel matrix-based simulations can be performed with the $hp$-infrastructure. In the present
release, we have extended the setup routines of \texttt{MatrixFree} so that they now also
provide parallel $hp$-support.

Until now, the \texttt{FEEvaluation} classes used the template parameters $k$ and $q$ to select the correct active FE and quadrature index and users had
to use the function \texttt{MatrixFree::create\_cell\_\allowbreak subrange\_\allowbreak hp()} or
\texttt{::create\_cell\_\allowbreak subrange\_\allowbreak hp\_\allowbreak by\_index()} to create subranges of cells with
the same polynomial degree. This led to user codes that were hard to read due to complicated jump tables.

The creation of subranges is now performed internally, and the non-templated versions
of the \texttt{FEEvaluation} classes are extended for the $hp$-case. To nevertheless determine
the active FE and quadrature index, the current cell/face range has to be provided
to the constructors of the \texttt{FEEvaluation} classes, from which the relevant information
can be deduced (in the simplex case also the face type). These changes enable
users to write matrix-free code independently of whether $hp$-capabilities are used or not.

@@

Example: testing of surface tension in the context of sharp-interface methods:
\begin{align*}
\left(\vec{v}, \kappa \, \vec{n}\right)_\Gamma
\approx
\sum\left(\vec{v}, \kappa(\vec{p}_q) \, \vec{n}(\vec{p}_q) (JxW)_q\right)
\end{align*}
in code:
\begin{c++}
phi_normal.evaluate(cell, points, normal_values, EvaluationFlags::values);
for (unsigned int q = 0; q < n_points; ++q)
   phi_force.submit_value(phi_normal.get_value(q) *
                          phi_curvature.get_value(q) * JxW[q], q);
phi_force.integrate(cell, points, force_values, EvaluationFlags::values);
\end{c++}
The quadrature points and the connected \texttt{JxW} can, e.g., come from a
codim-1 mesh. Determining to which \texttt{cell} a quadrature point belongs
to on the background mesh, incl. the reference-cell coordinates \texttt{points},
can be determined with know functions like \texttt{GridTools::find\_all\_active\_cells\_around\_point()}.

>>>>

\author[4]{Timo~Heister}
 \affil[4]{School of Mathematical and Statistical Sciences,
   Clemson University,
   Clemson, SC, 29634, USA
   {\texttt{tcleven/heister@clemson.edu}}}

\author[5]{Luca~Heltai}
\affil[5]{SISSA,
   International School for Advanced Studies,
   Via Bonomea 265,
   34136, Trieste, Italy.
   {\texttt{luca.heltai@sissa.it}}}

 \author[6,7]{Martin~Kronbichler}
 \affil[6]{Institute for Computational Mechanics,
   Technical University of Munich,
   Boltzmannstr.~15, 85748 Garching, Germany.
   {\texttt{kronbichler/munch@lnm.mw.tum.de}}}
 \affil[7]{Department of Information Technology,
   Uppsala University,
   Box 337, 751\,05 Uppsala, Sweden.
   {\texttt{martin.kronbichler@it.uu.se}}}

\author[8]{Matthias~Maier}
\affil[8]{Department of Mathematics,
  Texas A\&M University,
  3368 TAMU,
  College Station, TX 77845, USA.
  {\texttt{maier@math.tamu.edu}}}

\author[6,9]{Peter Munch}
 \affil[10]{Institute of Materials Research, Materials Mechanics,
 Helmholtz-Zentrum Geesthacht,
 Max-Planck-Str. 1, 21502 Geesthacht, Germany.
   {\texttt{peter.muench@hzg.de}}}

\author[10]{Jean-Paul~Pelteret}
\affil[10]{Independent researcher.
{\texttt{jppelteret@gmail.com}}}

@@

We provide matrix-free support for simplex and mixed meshes as well as for
continuous and discontinuous elements. From a user perspective, the main
changes are to pass a d-dimensional quadrature rules rather than a 1D formula,
and the fact that \texttt{FEEvaluation} and \texttt{FEFaceEvaluation} must not
specify the polynomial degree via template arguments, determining all
information at runtime. More details can be found in
Subsection~\ref{subsec:mf}.

By the time of writing, no advanced algorithms for evaluating values and
gradients at the quadrature points such as sum factorization are used.
The use of full interpolation matrices is for now acceptable since only low-order elements are
supported.

@@

Until now, \dealii has only supported local smoothing multigrid
algorithms~\citep{ClevengerHeisterKanschatKronbichler2019}; in local
smoothing algorithms, smoothers only act on the cells of a given refinement level,
skipping those parts of the mesh that are not adaptively refined to that level.
The current release now also has support for global coarsening~\citep{becker00} when
using continuous (\texttt{FE\_Q}, \texttt{FE\_SimplexP}) and
discontinuous (\texttt{FE\_DGQ}, \texttt{FE\_SimplexDGP})
elements. Global coarsening builds multigrid levels for the entire
domain, coarsening \textit{all} cells of a triangulation regardless of
how many times they have been refined. In addition, the framework now available in
\dealii{} in not only applicable to geometric coarsening, but can also
perform coarsening by reducing the polynomial degree ($p$-multigrid,
see~\cite{Fehn2020}), for example
to support $hp$-adaptive meshes.
%where it has been shown that one can
%efficiently precondition high-order spaces with lower-order finite
%element operators.
% MK: I do not think "precondition" is a precise enough description of p-multigrid
% because one still needs smoothing on the fine levels in addition to the
% "solve" at the coarser degree, a fact not transported by the mere
% 'precondition' term; rather than trying to discuss this in a more precise way,
% let us simply skip it as that is not central to the presentation here.
Finally, the implementation also supports transfer between continuous
and discontinuous elements as a further way to create multigrid levels.

@@

The transfer operators between two levels has been implemented in the new class \texttt{MGTwoLevelTransfer}, which can be set up via the functions \texttt{MGTwoLevelTransfer::\allowbreak reinit\_\allowbreak geometric\_\allowbreak transfer()} or \texttt{MGTwo\allowbreak LevelTransfer::\allowbreak reinit\_\allowbreak polynomial\_\allowbreak transfer()} for given
\texttt{DoFHandler} and \texttt{AffineConstraint} classes of two levels. The resulting transfer operators
can then be collected in a single
\texttt{MGTransferGlobalCoarsening} object that can be used just as the previous workhorse \texttt{MGTransferMatrixFree} within the \texttt{Multigrid}
algorithm.
%Several common operations are encoded in utility
%functions in the \texttt{MGTransferGlobalCoarseningTools}
%namespace.
To facilitate the construction of matrix diagonals with matrix-free methods as well as a matrix representation of the coarse level matrix, new utility functions \texttt{create\_diagonal()} and \texttt{create\_matrix()} from
the \texttt{MatrixFreeTools} namespace have been added. (see also Subsection~\ref{subsec:mf}).

@@

The classes \texttt{FEEvaluation} and \texttt{FEFaceEvaluation} use template parameters for
the polynomial degree of the finite element $k$ and the number of the 1D quadrature points $q$
to generate near-optimal code for these operations.
For application codes that rely on operators of many different degrees (e.g., because
they use $p$-multigrid or $hp$-algorithms), creating all instantiations
can be overly complex and incur long compile times.

In the current release, specializations of these classes that do
not rely on the template parameters $k$ and $q$ (expressed in the code
with ``-1'' and ``0'') have been added. For example:
\begin{c++}
FEEvaluation<dim, -1, 0, n_components, Number, VectorizedArrayType>
  phi(range, dofhandler_index, quadrature_index, first_selected_component);
\end{c++}
These classes select at runtime -- for common low and medium
polynomial degree/quadrature combinations ($k\le 6$ and $q\in\{ k+1,
k+2, \left\lfloor (3k)/2 \right\rfloor \}$) -- efficient precompiled implementations and default to non-templated
evaluation kernels else.

@@

With release 9.1 \cite{dealII91}, large parts of the $hp$-algorithms in \dealii were parallelized, enabling
parallel matrix-based simulations with the $hp$-infrastructure. In the present
release, the parallel $hp$ support was extended to \texttt{MatrixFree}.

Until now, the \texttt{FEEvaluation} classes used the template parameters $k$ and $q$ to select the correct active FE and quadrature index and cumbersome detection of subranges of the same degree.
The creation of subranges is now performed internally, and the non-templated versions
of the \texttt{FEEvaluation} classes have been extended for the $hp$-case. To determine
the desired FE and quadrature index of a subrange, the current cell/face range has to be provided
to the constructors of the \texttt{FEEvaluation} classes. These changes enable
users to write matrix-free code independently of whether $hp$-capabilities are used or not.

@@

In a number of circumstances, finite element solutions need to be evaluated on
arbitrary reference points that change from one element to next. Two important
examples are particle simulations coupled to a finite element solution, or
algorithms on non-matching grids. The existing \texttt{FEValues} class a poor
fit for this task, as it relies on tabulating all information before use,
necessitating a separate setup for every cell. The new class
\texttt{FEPointEvaluation} provides a more convenient interface for this
task. For tensor product finite elements (\texttt{FE\_Q}, \texttt{FE\_DGQ})
and tensor product mappings (\texttt{MappingQGeneric} and derived classes),
the new approach is also very fast, as it can use some of the matrix-free
infrastructure and vectorization facilities.

To give an example of the new features, let us consider the evaluation of a
surface tension force in the context of sharp-interface methods, whose
contribution is added to a fluid solver by multiplication with test function
and addition over quadrature points:
\begin{align*}
\left(\vec{v}, \kappa \, \vec{n}\right)_\Gamma
\approx
\sum_q\left(\vec{v}, \kappa(\vec{p}_q) \, \vec{n}(\vec{p}_q) (JxW)_q\right).
\end{align*}
In \dealii{}, this can now be conveniently written as
\begin{c++}
phi_normal.reinit(cell, reference_points);
phi_force.reinit(cell, reference_points);
phi_normal.evaluate(normal_values, EvaluationFlags::values);
for (unsigned int q = 0; q < n_points; ++q)
   phi_force.submit_value(phi_normal.get_value(q) *
                          phi_curvature.get_value(q) * JxW[q], q);
phi_force.integrate(force_values, EvaluationFlags::values);
\end{c++}
The quadrature points and the related \texttt{JxW} value can, e.g., come from
a mesh of codimension one. Determining to which \texttt{cell} a quadrature
point belongs to on the background mesh, including the reference-cell
coordinates \texttt{reference\_points}, can be determined with functions like
\texttt{GridTools::find\_all\_active\_cells\_around\_point()}. While these
functions have been available in \dealii{} previously, their performance has
been considerably enhanced with the aforementioned more optimized code paths
for selected mappings.
